
# Overview

This R Markdown file prepares imaging mass cytometry (IMC) breast cancer data for downstream spatial and clinical analysis. The main steps are:

1. Inspect source RDS/Seurat objects.
2. Prepare patient-level metadata.
3. Prepare spatial coordinates and cluster labels.
4. Prepare protein-expression matrices.
5. Generate per-core expression and spatial files.
6. Perform basic sample-size quality control.
7. Assemble cleaned clinical metadata.

> **Note:** Update the paths in the configuration section before running this notebook. Raw data files are not included in the repository.

# 1. Setup

```{r setup, message=FALSE, warning=FALSE}
library(Seurat)
library(dplyr)
library(readr)
library(ggplot2)
library(tidyverse)
library(tools)

# Optional packages used by downstream analyses.
# Uncomment if needed and already installed.
# library(PQLseq)

# -----------------------------------------------------------------------------
# User configuration
# -----------------------------------------------------------------------------
# Recommended: set the BSNMANI_BC_DIR environment variable on your machine.
# Example Windows value: C:/Users/<username>/Desktop/BSNMANi/BC
base_dir <- Sys.getenv("BSNMANI_BC_DIR", unset = "path/to/BSNMANi/BC")

# A second metadata location used near the end of the notebook.
tidybc_dir <- Sys.getenv(
  "BSNMANI_TIDYBC_DIR",
  unset = "path/to/BSNMANi/TidyBC/Shu_BaselTMA"
)

# Non-protein morphological/features columns to remove from expression matrices.
non_protein_features <- c(
  "Area", "Eccentricity", "EulerNumber", "Extent",
  "MajorAxisLength", "MinorAxisLength", "Number_Neighbors",
  "Orientation", "Percent_Touching", "Perimeter",
  "phospho", "Solidity"
)
```

# 2. Inspect the Seurat / Patient Data Object

```{r inspect-patient-data}
patient_data_path <- file.path(base_dir, "patient_data.rds")
patient_obj <- readRDS(patient_data_path)

str(patient_obj)
head(patient_obj)
```

## 2.1 Sort patient data by the numeric suffix of the row ID

```{r sort-patient-data}
patient_df <- patient_obj

# Convert row names to an explicit ID column.
patient_df$id <- rownames(patient_df)

# Extract the trailing numeric component of each ID and sort by it.
patient_df$order <- as.numeric(sub(".*_", "", patient_df$id))
patient_df <- patient_df[order(patient_df$order), ]
patient_df$order <- NULL

# Move the ID column to the first position.
patient_df <- patient_df[, c("id", setdiff(colnames(patient_df), "id"))]

# Round all non-ID columns after coercing them to numeric values.
patient_df[, -1] <- lapply(patient_df[, -1, drop = FALSE], function(x) {
  round(as.numeric(x))
})

write.csv(
  patient_df,
  file.path(base_dir, "patient_data_sorted_indexed.csv"),
  row.names = FALSE
)
```

## 2.2 Extract Seurat assay matrices

```{r extract-seurat-assays}
# Count matrix from the default assay, typically RNA.
count_mat <- GetAssayData(patient_obj, slot = "counts")

# Log-normalized expression matrix.
norm_mat <- GetAssayData(patient_obj, slot = "data")

# Explicitly extract RNA counts when multiple assays are present.
count_mat_rna <- GetAssayData(patient_obj, assay = "RNA", slot = "counts")
```

# 3. Inspect Spatial and Metal-Count Inputs

## 3.1 Read a spatial CSV file

```{r inspect-spatial-csv}
patient_spatial_path <- file.path(base_dir, "patient_spatial.csv")
patient_spatial <- read.csv(patient_spatial_path, header = TRUE, row.names = 1)
head(patient_spatial)
```

## 3.2 Inspect the Basel metal-count RDS object

```{r inspect-metal-count-rds}
metal_count_path <- file.path(base_dir, "Basel_patient_metal_count.RDS")
metal_count_obj <- readRDS(metal_count_path)

class(metal_count_obj)
str(metal_count_obj, max.level = 2)

if (is.data.frame(metal_count_obj) || is.matrix(metal_count_obj)) {
  head(metal_count_obj)
}

# Example core used for manual inspection.
example_core <- "BaselTMA_SP41_257_X3Y1"
if (example_core %in% names(metal_count_obj)) {
  example_core_data <- metal_count_obj[[example_core]]
  print(dim(example_core_data))
  print(example_core_data[1:min(5, nrow(example_core_data)),
                          1:min(5, ncol(example_core_data))])
}
```

# 4. Prepare Spatial Information for One Patient/Core

```{r prepare-single-spatial-info}
spatial <- read.csv(file.path(base_dir, "patient_spatial.csv"))
clusters <- read.csv(
  file.path(
    base_dir,
    "singlecell_cluster_labels",
    "Cluster_labels",
    "Basel_metaclusters.csv"
  )
)

# Keep the original spatial coordinates without rounding.
spatial_out <- data.frame(
  id = spatial$id,
  X = spatial$Location_Center_X,
  Y = spatial$Location_Center_Y
)

# Add cluster labels.
merged_spatial <- merge(
  spatial_out,
  clusters[, c("id", "cluster")],
  by = "id"
)

# Restore the original spatial-cell ordering.
merged_spatial$id <- factor(merged_spatial$id, levels = spatial$id)
merged_spatial <- merged_spatial[order(merged_spatial$id), ]
merged_spatial$id <- as.character(merged_spatial$id)

write.csv(
  merged_spatial,
  file.path(base_dir, "spatial_info.csv"),
  row.names = FALSE
)
```

# 5. Prepare a Protein-Expression Matrix for One Patient/Core

```{r prepare-single-protein-matrix}
patient_expr <- read.csv(file.path(base_dir, "patient_data_sorted_indexed.csv"))

# Identify non-marker columns. Here, terms such as "phospho" also match columns
# whose names contain that term.
non_marker_cols <- unique(unlist(lapply(
  non_protein_features,
  function(x) grep(x, colnames(patient_expr), value = TRUE)
)))

# Retain ID plus protein-marker columns.
marker_cols <- setdiff(colnames(patient_expr), non_marker_cols)
marker_cols <- union("id", setdiff(marker_cols, "id"))
protein_expr <- patient_expr[, marker_cols]

write.csv(
  protein_expr,
  file.path(base_dir, "protein_expression_matrix.csv"),
  row.names = FALSE
)
```

# 6. Prepare Spatial Information for All Cores

```{r prepare-all-spatial-info}
locations <- read.csv(
  file.path(base_dir, "singlecell_locations", "Basel_SC_locations.csv")
)
clusters <- read.csv(
  file.path(
    base_dir,
    "singlecell_cluster_labels",
    "Cluster_labels",
    "Basel_metaclusters.csv"
  )
)

spatial_all <- data.frame(
  id = locations$id,
  X = locations$Location_Center_X,
  Y = locations$Location_Center_Y,
  core = locations$core
)

# Keep all cells from the spatial-location table.
spatial_all <- merge(
  spatial_all,
  clusters[, c("id", "cluster")],
  by = "id",
  all.x = TRUE
)

# Label cells without matched cluster annotations.
spatial_all$cluster[is.na(spatial_all$cluster)] <- "unknown"

# Restore the original cell order.
spatial_all$id <- factor(spatial_all$id, levels = locations$id)
spatial_all <- spatial_all[order(spatial_all$id), ]
spatial_all$id <- as.character(spatial_all$id)

write.csv(
  spatial_all,
  file.path(base_dir, "spatial_info_all.csv"),
  row.names = FALSE
)
```

## 6.1 Export one spatial CSV per core

```{r export-spatial-by-core}
spatial_output_dir <- file.path(base_dir, "spatial_data")
dir.create(spatial_output_dir, recursive = TRUE, showWarnings = FALSE)

for (core_name in unique(spatial_all$core)) {
  core_df <- spatial_all[
    spatial_all$core == core_name,
    c("id", "X", "Y", "cluster")
  ]

  # Sort by the trailing numeric cell ID when available.
  cell_number <- suppressWarnings(
    as.numeric(gsub(".*_(\\d+)$", "\\1", core_df$id))
  )
  core_df <- core_df[order(cell_number, na.last = TRUE), ]

  safe_name <- gsub("[^A-Za-z0-9_]", "_", core_name)
  write.csv(
    core_df,
    file.path(spatial_output_dir, paste0(safe_name, ".csv")),
    row.names = FALSE
  )
}
```

## 6.2 Inspect the cluster distribution in one core

```{r inspect-core-clusters}
example_spatial_file <- file.path(
  spatial_output_dir,
  "BaselTMA_SP41_101_X10Y8.csv"
)

if (file.exists(example_spatial_file)) {
  example_spatial_df <- read.csv(example_spatial_file)
  print(table(example_spatial_df$cluster))
}
```

# 7. Inspect Clinical Data

```{r inspect-clinical-data}
clinical_path <- file.path(base_dir, "clinical_data.RDS")
clinical_obj <- readRDS(clinical_path)

class(clinical_obj)
str(clinical_obj)
names(clinical_obj)

# Retain patients with available survival status and overall-survival time.
surv_data <- clinical_obj[
  !is.na(clinical_obj$Patientstatus) & !is.na(clinical_obj$OSmonth),
]
cat("Number of patients with survival information:", nrow(surv_data), "\n")

print(table(clinical_obj$Patientstatus))

alive_data <- clinical_obj[
  !is.na(clinical_obj$Patientstatus) &
    !is.na(clinical_obj$OSmonth) &
    clinical_obj$Patientstatus == 0,
]
cat(
  "Number of patients with survival information and status = 0:",
  nrow(alive_data),
  "\n"
)
```

# 8. Expression-Matrix Helper Functions

The original notebook repeated the same transpose, ID-recovery, sorting, and filtering operations for several output variants. The following helper functions centralize those operations without changing their intended behavior.

```{r expression-helper-functions}
extract_expression_component <- function(core_object, prefer_count = FALSE) {
  if (
    prefer_count &&
    is.list(core_object) &&
    !is.data.frame(core_object) &&
    "count" %in% names(core_object)
  ) {
    return(core_object$count)
  }
  core_object
}

format_core_expression <- function(
  expr_mat,
  excluded_columns = non_protein_features
) {
  # Transpose to cells x proteins/features.
  expr_t <- as.data.frame(t(expr_mat), stringsAsFactors = FALSE)
  cell_ids <- rownames(expr_t)

  if (nrow(expr_t) < 2) {
    stop("The transposed expression matrix has fewer than two rows.")
  }

  # Preserve the layout used in the original preprocessing notebook:
  # the first transposed row stores the output column names.
  colnames(expr_t) <- as.character(expr_t[1, ])
  expr_t <- expr_t[-1, , drop = FALSE]

  # Add the original cell IDs back as the first column.
  expr_t <- cbind(id = cell_ids[-1], expr_t, stringsAsFactors = FALSE)

  # Sort cells by the trailing numeric component of the ID.
  id_numbers <- suppressWarnings(
    as.numeric(sub(".*_(\\d+)$", "\\1", expr_t$id))
  )
  expr_t <- expr_t[order(id_numbers, na.last = TRUE), , drop = FALSE]
  rownames(expr_t) <- NULL

  # Remove non-protein columns using exact column-name matching, consistent
  # with the repeated batch-processing blocks in the original notebook.
  expr_t <- expr_t[, !(colnames(expr_t) %in% excluded_columns), drop = FALSE]
  expr_t
}

coerce_expression_to_numeric <- function(expr_df) {
  numeric_cols <- setdiff(colnames(expr_df), "id")
  expr_df[, numeric_cols] <- lapply(
    expr_df[, numeric_cols, drop = FALSE],
    function(x) suppressWarnings(as.numeric(as.character(x)))
  )
  expr_df
}
```

# 9. Export Expression Matrices for All Cores

Reload the metal-count object here if this section is run independently.

```{r reload-metal-count-object}
metal_count_obj <- readRDS(file.path(base_dir, "Basel_patient_metal_count.RDS"))
```

## 9.1 Raw filtered expression matrices

```{r export-raw-expression}
raw_expr_dir <- file.path(base_dir, "expression_matrix")
dir.create(raw_expr_dir, recursive = TRUE, showWarnings = FALSE)

for (core_name in names(metal_count_obj)) {
  core_object <- metal_count_obj[[core_name]]
  expr_mat <- extract_expression_component(core_object, prefer_count = TRUE)
  expr_t <- format_core_expression(expr_mat)

  safe_name <- gsub("[^A-Za-z0-9_]", "_", core_name)
  write.csv(
    expr_t,
    file.path(raw_expr_dir, paste0(safe_name, ".csv")),
    row.names = FALSE
  )
}
```

## 9.2 Rounded expression matrices

```{r export-rounded-expression}
rounded_expr_dir <- file.path(base_dir, "expression_matrix_round")
dir.create(rounded_expr_dir, recursive = TRUE, showWarnings = FALSE)

for (core_name in names(metal_count_obj)) {
  core_object <- metal_count_obj[[core_name]]
  expr_mat <- extract_expression_component(core_object, prefer_count = TRUE)
  expr_t <- format_core_expression(expr_mat)
  expr_t <- coerce_expression_to_numeric(expr_t)

  numeric_cols <- setdiff(colnames(expr_t), "id")
  expr_t[, numeric_cols] <- lapply(
    expr_t[, numeric_cols, drop = FALSE],
    round
  )

  if (any(unlist(expr_t[, numeric_cols, drop = FALSE]) < 0, na.rm = TRUE)) {
    message(sprintf(
      "Warning: negative values detected after rounding in core '%s'.",
      core_name
    ))
  }

  safe_name <- gsub("[^A-Za-z0-9_]", "_", core_name)
  write.csv(
    expr_t,
    file.path(rounded_expr_dir, paste0(safe_name, ".csv")),
    row.names = FALSE
  )
}
```

## 9.3 Row-normalized expression matrices scaled to 10,000

```{r export-scaled-expression}
scaled_expr_dir <- file.path(base_dir, "expression_matrix_scaled10000")
dir.create(scaled_expr_dir, recursive = TRUE, showWarnings = FALSE)

for (core_name in names(metal_count_obj)) {
  core_object <- metal_count_obj[[core_name]]
  expr_mat <- extract_expression_component(core_object, prefer_count = TRUE)
  expr_t <- format_core_expression(expr_mat)
  expr_t <- coerce_expression_to_numeric(expr_t)

  numeric_cols <- setdiff(colnames(expr_t), "id")
  numeric_matrix <- as.matrix(expr_t[, numeric_cols, drop = FALSE])
  storage.mode(numeric_matrix) <- "numeric"

  if (!is.numeric(numeric_matrix)) {
    stop("The expression matrix still contains non-numeric values.")
  }

  # Normalize each cell by its total expression and scale to 10,000.
  row_sums <- rowSums(numeric_matrix, na.rm = TRUE)
  row_sums[row_sums == 0] <- 1

  expr_norm <- sweep(numeric_matrix, 1, row_sums, FUN = "/")
  expr_scaled <- round(expr_norm * 10000)
  expr_t[, numeric_cols] <- as.data.frame(expr_scaled)

  if (any(unlist(expr_t[, numeric_cols, drop = FALSE]) < 0, na.rm = TRUE)) {
    message(sprintf(
      "Warning: negative values detected after normalization in core '%s'.",
      core_name
    ))
  }

  safe_name <- gsub("[^A-Za-z0-9_]", "_", core_name)
  write.csv(
    expr_t,
    file.path(scaled_expr_dir, paste0(safe_name, ".csv")),
    row.names = FALSE
  )
}
```

# 10. Spatial-File Cell-Count Quality Control

```{r spatial-file-qc}
spatial_csv_files <- list.files(
  spatial_output_dir,
  pattern = "\\.csv$",
  full.names = TRUE
)

num_rows <- sapply(spatial_csv_files, function(file) {
  df <- tryCatch(read.csv(file), error = function(e) NULL)
  if (is.data.frame(df)) nrow(df) else NA_integer_
})
num_rows <- num_rows[!is.na(num_rows)]

cat("Minimum number of rows:", min(num_rows), "\n")
cat("Maximum number of rows:", max(num_rows), "\n")

spatial_count_df <- data.frame(rows = num_rows)
ggplot(spatial_count_df, aes(x = rows)) +
  geom_histogram(binwidth = 10) +
  theme_minimal() +
  labs(
    title = "Distribution of Cell Counts Across Spatial CSV Files",
    x = "Number of cells",
    y = "Number of files"
  )
```

## 10.1 Detect low-count outliers in log space

```{r spatial-file-outliers}
log_rows <- log1p(num_rows)
Q1 <- quantile(log_rows, 0.25)
Q3 <- quantile(log_rows, 0.75)
IQR_value <- Q3 - Q1
lower_bound <- Q1 - 1.5 * IQR_value

outlier_idx <- which(log_rows < lower_bound)
outlier_info <- data.frame(
  file_name = basename(names(num_rows[outlier_idx])),
  row_count = num_rows[outlier_idx]
)

cat("Files below the lower IQR bound:\n")
print(outlier_info)

log_count_df <- data.frame(log_rows = log_rows)
ggplot(log_count_df, aes(y = log_rows)) +
  geom_boxplot() +
  labs(
    title = "Boxplot of log(Number of Cells + 1)",
    y = "log1p(Number of cells)"
  ) +
  theme_minimal()
```

## 10.2 Summarize retained spatial files

```{r summarize-retained-spatial-files}
filtered_rows <- if (length(outlier_idx) > 0) {
  num_rows[-outlier_idx]
} else {
  num_rows
}

cat("Number of files after outlier removal:", length(filtered_rows), "\n")
cat("Mean number of cells:", mean(filtered_rows), "\n")
cat("Median number of cells:", median(filtered_rows), "\n")

filtered_count_df <- data.frame(rows = filtered_rows)
ggplot(filtered_count_df, aes(x = rows)) +
  geom_histogram(binwidth = 50) +
  theme_minimal() +
  labs(
    title = "Distribution of Cell Counts After Outlier Removal",
    x = "Number of cells",
    y = "Number of files"
  )

retained_info <- data.frame(
  file_name = basename(names(filtered_rows)),
  row_count = as.integer(filtered_rows)
)
retained_info <- retained_info[order(retained_info$row_count), ]
print(head(retained_info, 10))
```

# 11. Assemble Clean Clinical Metadata

Use the spatial cores retained after the spatial-file quality-control step to assemble the corresponding clinical metadata.

```{r assemble-clinical-metadata}
metadata_path <- file.path(tidybc_dir, "Basel_PatientMetadata.csv")
metadata_df <- read_csv(metadata_path, show_col_types = FALSE)

# Convert retained spatial CSV filenames to core names.
retained_core_df <- retained_info %>%
  mutate(core = file_path_sans_ext(file_name)) %>%
  select(core)

clinical_df <- retained_core_df %>%
  left_join(
    metadata_df %>%
      select(core, age, OSmonth, Patientstatus, grade, clinical_type),
    by = "core"
  ) %>%
  filter(!is.na(age) & !is.na(OSmonth))

print(clinical_df)

saveRDS(
  clinical_df,
  file = file.path(base_dir, "clinical_df_3cov.RDS")
)
```
