---
title: "Co-expression Network Matrix Assembly"
output: html_document
---

# Overview

This notebook contains the co-expression-network-related portion separated from the original preprocessing workflow.

# 1. Setup

```{r setup, message=FALSE, warning=FALSE}
library(readr)
library(dplyr)
library(tools)

# -----------------------------------------------------------------------------
# User configuration
# -----------------------------------------------------------------------------
# Recommended: define BSNMANI_BC_DIR on your local system.
# Example:
#   Sys.setenv(BSNMANI_BC_DIR = "path/to/BSNMANi/BC")
base_dir <- Sys.getenv(
  "BSNMANI_BC_DIR",
  unset = "path/to/BSNMANi/BC"
)

# Metadata containing the cores to retain and their desired order.
core_metadata_file <- file.path(base_dir, "Basel_CleanCore_Age.csv")

# Folder containing precomputed per-core correlation matrices.
correlation_dir <- file.path(base_dir, "correlation_matrix")

# Output file.
output_rds <- file.path(base_dir, "WGCNA_transformed_BC.RDS")
```

# 2. Load the Core List

```{r load-core-list}
if (!file.exists(core_metadata_file)) {
  stop("Core metadata file not found: ", core_metadata_file)
}

core_metadata <- read_csv(
  core_metadata_file,
  show_col_types = FALSE
)

if (!"core" %in% colnames(core_metadata)) {
  stop("The metadata file must contain a column named 'core'.")
}

core_list <- unique(core_metadata$core)

cat("Number of cores requested:", length(core_list), "\n")
head(core_list)
```

# 3. Load Precomputed Correlation Matrices

Each matrix is expected to be stored as:

```text
correlation_matrix/
├── corr_<core_1>.csv
├── corr_<core_2>.csv
├── corr_<core_3>.csv
└── ...
```

```{r load-correlation-matrices}
correlation_list <- list()
missing_cores <- character()

for (core_name in core_list) {

  file_name <- paste0("corr_", core_name, ".csv")
  file_path <- file.path(correlation_dir, file_name)

  if (!file.exists(file_path)) {
    warning("Correlation matrix not found: ", file_name)
    missing_cores <- c(missing_cores, core_name)
    next
  }

  correlation_matrix <- as.matrix(
    read.csv(
      file_path,
      row.names = 1,
      check.names = FALSE
    )
  )

  # Ensure that matrix values are numeric.
  storage.mode(correlation_matrix) <- "numeric"

  correlation_list[[core_name]] <- correlation_matrix
}

cat(
  "Successfully loaded:",
  length(correlation_list),
  "of",
  length(core_list),
  "correlation matrices.\n"
)

if (length(missing_cores) > 0) {
  cat("Number of missing cores:", length(missing_cores), "\n")
  print(missing_cores)
}
```

# 4. Quality Control

The downstream network representation should contain one square marker-by-marker matrix for each core.

```{r correlation-matrix-qc}
validate_correlation_matrix <- function(mat, core_name, tolerance = 1e-8) {

  issues <- character()

  if (!is.matrix(mat)) {
    issues <- c(issues, "not a matrix")
    return(issues)
  }

  if (nrow(mat) != ncol(mat)) {
    issues <- c(issues, "matrix is not square")
  }

  if (anyNA(mat)) {
    issues <- c(issues, "contains missing values")
  }

  if (nrow(mat) == ncol(mat)) {
    if (!isTRUE(all.equal(
      mat,
      t(mat),
      tolerance = tolerance,
      check.attributes = FALSE
    ))) {
      issues <- c(issues, "matrix is not symmetric")
    }
  }

  if (!is.null(rownames(mat)) &&
      !is.null(colnames(mat)) &&
      !identical(rownames(mat), colnames(mat))) {
    issues <- c(
      issues,
      "row and column marker names are not identically ordered"
    )
  }

  if (length(issues) == 0) {
    return("OK")
  }

  paste(issues, collapse = "; ")
}

qc_results <- data.frame(
  core = names(correlation_list),
  n_features = vapply(
    correlation_list,
    nrow,
    integer(1)
  ),
  status = vapply(
    names(correlation_list),
    function(core_name) {
      validate_correlation_matrix(
        correlation_list[[core_name]],
        core_name
      )
    },
    character(1)
  ),
  stringsAsFactors = FALSE
)

print(qc_results)

problematic_matrices <- qc_results %>%
  filter(status != "OK")

if (nrow(problematic_matrices) > 0) {
  warning(
    nrow(problematic_matrices),
    " correlation matrices did not pass all QC checks."
  )
  print(problematic_matrices)
}
```

# 5. Check Marker Consistency Across Cores

All cores should ideally use the same markers in the same order before the matrices are passed to downstream modeling.

```{r marker-consistency}
if (length(correlation_list) > 0) {

  reference_core <- names(correlation_list)[1]
  reference_markers <- rownames(correlation_list[[reference_core]])

  inconsistent_cores <- names(correlation_list)[
    !vapply(
      correlation_list,
      function(mat) {
        identical(rownames(mat), reference_markers) &&
          identical(colnames(mat), reference_markers)
      },
      logical(1)
    )
  ]

  if (length(inconsistent_cores) == 0) {
    cat("All correlation matrices use the same marker order.\n")
  } else {
    warning(
      "Marker names/order differ for ",
      length(inconsistent_cores),
      " core(s)."
    )
    print(inconsistent_cores)
  }
}
```

# 6. Assemble the Network-Matrix List

The original preprocessing notebook named this object `WGCNA_transformed_BC`.

```{r assemble-network-list}
# Keep the matrices in the same order as core_list.
available_cores <- core_list[core_list %in% names(correlation_list)]

WGCNA_transformed_BC <- correlation_list[available_cores]

cat(
  "Final number of co-expression matrices:",
  length(WGCNA_transformed_BC),
  "\n"
)

head(names(WGCNA_transformed_BC))
```

# 7. Save the RDS Object

```{r save-network-rds}
saveRDS(
  WGCNA_transformed_BC,
  file = output_rds
)

cat("Saved network matrices to:\n", output_rds, "\n")
```

# 8. Inspect the Final Object

```{r inspect-network-object}
network_obj <- readRDS(output_rds)

cat("Object class:", class(network_obj), "\n")
cat("Number of cores:", length(network_obj), "\n")

if (length(network_obj) > 0) {

  first_core <- names(network_obj)[1]

  cat("Example core:", first_core, "\n")
  cat(
    "Matrix dimensions:",
    paste(dim(network_obj[[first_core]]), collapse = " x "),
    "\n"
  )

  print(
    network_obj[[first_core]][
      1:min(5, nrow(network_obj[[first_core]])),
      1:min(5, ncol(network_obj[[first_core]])),
      drop = FALSE
    ]
  )
}
```
